Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-12 intro.tex
changed fig 1 name; rewrote section 3.1

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  While this is not a common laboratory
experiment, it is a highly appropriate choice here.  Computers are
among the most complex engineered artifacts in current use.  Modern
microprocessor chips contain multiple processing units and multi-layer
memories, for instance, and they use complicated hardware/software
strategies to move data and threads of computation across those
resources.  As a result, the processor and memory loads during the
execution of even a very simple program can evolve in a very
complicated fashion.  Figure~\ref{fig:ipc} shows a performance trace
of a three-line loop that repeatedly initializes the upper triangle of
a matrix in column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the program can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate the
  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

>>>>

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  While this is not a common laboratory
experiment, it is a highly appropriate choice here.  Computers are
among the most complex engineered artifacts in current use.  Modern
microprocessor chips contain multiple processing units and multi-layer
memories, for instance, and they use complicated hardware/software
strategies to move data and threads of computation across those
resources.  As a result, the processor and memory loads during the
execution of even a very simple program can evolve in a very
complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance trace
of a three-line loop that repeatedly initializes the upper triangle of
a matrix in column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the program can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate the
  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

### 2014-03-12 methods.tex
changed fig 1 name; rewrote section 3.1

The time-series data for these experiments was collected on an Intel Core\textsuperscript{\textregistered} i7-2600 running the 2.6.38-8 Linux
kernel.  This Nehalem chip has eight cores running at 3.40GHz and a cache size
of 8192 KB.  We collected performance traces of processor efficiency (IPC) during the execution of three exemplar programs---a simple microkernel(\col) and two complex programs: one from the
SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd). In particular, we measured and aggregated the instructions executed per clock cycle (ipc)---a measure of processor efficiency---at 100,000-instruction intervals.  To record these measurements,
we used the {\tt libpfm4} library, via PAPI 5.2
\footnote{Performance Application Programming Interface}\cite{papi}, combined with a custom measurement infrastructure\cite{todd-phd} to stop program execution every 100,00-instructions and read the contents of the CPU's onboard hardware performance monitors (HPMs). HPMs are specialty registers who exist solely to store performance data.  For statistical validation we collected 15 performance traces from each program resulting in 45 total time series.

Obviously using a system to measure itself can cause interference--i.e, are you measuring the generating process cleanly or are you adding noise to the process?---but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way. For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see 
\cite{zach-IDA10,mytkowicz09,todd-phd}.

>>>>

The time-series data for these experiments was collected on an Intel
Core\textsuperscript{\textregistered} i7-2600-based machine running
the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
eight processing units---''cores''---a clock rate of 3.40GHz, and a
cache size of 8192 KB.  We collected performance traces of processor
efficiency during the execution of three different programs---the
simple loop (\col) whose performance is depicted in
Figure~\ref{fig:col-ipc} and two more-complex programs: one from the
SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK linear
algebra package (\svd).  To record these measurements, we used the
{\tt libpfm4} library, via PAPI
% 
% \footnote{Performance Application Programming Interface}
% 
5.2~\cite{papi}, to stop program execution at 100,000-instruction
intervals and read the contents of the CPU's onboard hardware
performance monitors\footnote{These specialty registers are built into
  modern microprocessors for the purpose of on-board monitoring and
  storage of performance data.}, which we had programmed to count how
many instructions were executed in each clock cycle (IPC).  For an
in-depth description of this custom-measurement infrastructure,
including discussion of the implications of the measurement interval,
please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
validation, we collected 15 performance traces from each of the three
programs.

% Obviously using a system to measure itself can cause
% interference--i.e, are you measuring the generating process cleanly or
% are you adding noise to the process?---but due diligence was put into
% monitoring the generating process of these time series without
% interfering with it in a significant way.
